### 2017-05-30 paper/sexSelectionSpacing-ver1.tex
Restructure and edit Introduction
- move paper purpose to later
- move paper purpose to separate paragraph
- add application to India with brief outline of 3 periods
- minor edits for first two paragraphs

Spacing between births has long been used as a measure of son preference 
\citep{Leung1988}.
Before prenatal sex determination became available, the only recourse for
parents who wanted a son, but did not yet have one, was to have the 
next birth sooner.
The result was shorter average spacing after the births of girls than boys,
and worse health outcomes for girls.%
\footnote{
See, for example, 
\citet{Das1987}, \citet{Rahman1993}, \citet{Pong1994}, \citet{Haughton1996},
\citet{Arnold1997}, and \citet{Soest2012} on shorter birth spacing after
girls
and 
\citet{arnold98}, \citet{Whitworth2002}, \citet{Rutstein2005},
and \citet{Conde-Agudelo2006} on the association between shorter birth
spacing and worse health outcomes for girls.
Parents were also more likely to cease childbearing after the birth of 
a son than after a daughter 
\citep{ben-porath76b,Das1987,Arnold1997,clark00,filmer09,Altindag2016}
}

What has not previously been appreciated is that the introduction 
of prenatal sex determination substantially changed the relationship between 
son preference and birth spacing.
The issue is that each sex-selective abortion increases the duration between 
observed births by approximately a year.%
\footnote{
This increase consists of three parts starting from the time of the 
abortion.
First, the uterus needs at least two menstrual cycles to recover 
before conception, because a less than three months space 
between abortion and conception substantially increases the likelihood 
of a spontaneous abortion \citep{zhou00b}.
Second, an average of 6 months to next conception \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.
The waiting time to conception does vary by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}
We therefore now have a situation where families with the \emph{strongest} son 
preference likely show the \emph{longest} spacing after the birth of a daughter.
But, to further complicate matters, we may still observe short spacing 
after the birth of daughter as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.
In this paper, I introduce and apply a novel method that
directly incorporates the effects of sex-selective abortions on
the likelihood of a son being born \emph{and} the duration between births.
The method can be used to analyse both situations with and without prenatal
sex selection.

How spacing interacts with sex selection is important for two reasons.
First, spacing, by itself, can no longer be used to measure son preference. 
Spacing is still important and useful in understanding son preference, but 
only if combined with the likelihood of observing a boy or a girl.
Short spacing after a girl is indicative of son preference if observed
births are close to the natural sex ratio, whereas long spacing is indicative
of son preference if observed births are male biased.

Second, the duration since last birth may affect the decision on sex selection.
For example, as the duration from last birth becomes long enough, parents 
may reverse their decision to use prenatal sex determination,
and carry the next pregnancy to term whether male or female.
Hence, by examining whether sex selection decisions change with
spacing, we can draw a more nuanced picture of the degree of son preference.

% [contribution of paper]
% [TK maybe not focus so much on the simple method but rather understanding son preference].
My main contribution here is the introduction and application
of a novel method that directly incorporates that sex-selective 
abortions affect both the likelihood of a son being born \emph{and} the 
duration between births.
[TK move to estimation strategy?]
The empirical model is a competing risk, non-proportional hazard model with 
two exit states: either a boy or a girl is born.

I use the method to show how spacing between births has changed with
the introduction prenatal sex determination and how birth spacing plays 
an important role in our understanding of sex selection decisions and 
son preferences.

% [Why India?]
I apply the method to birth histories from India's National Family and Health 
Surveys for Hindu women covering the period 1972 to 2006. 
India is a particularly interesting case.
On one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period \citep{Yi1993,park95}.
}
On the other hand, recent research suggests, however, that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
}

>>>>

Spacing between births has long been used as a measure of son preference 
\citep{Leung1988}.
Before prenatal sex determination became available, the only recourse for 
parents who wanted a son---but did not yet have one---was to have the next 
birth sooner.
Son preference was therefore often associated with shorter average spacing 
after the births of girls than boys, which, in turn, was associated with worse 
health outcomes for girls.%
\footnote{
See, for example, 
\citet{Das1987}, \citet{Rahman1993}, \citet{Pong1994}, \citet{Haughton1996},
\citet{Arnold1997}, and \citet{Soest2012} on shorter birth spacing after
girls
and 
\citet{arnold98}, \citet{Whitworth2002}, \citet{Rutstein2005},
and \citet{Conde-Agudelo2006} on the association between shorter birth
spacing and worse health outcomes for girls.
Parents were also more likely to cease childbearing after the birth of 
a son than after a daughter 
\citep{ben-porath76b,Das1987,Arnold1997,clark00,filmer09,Altindag2016}
}

What has not previously been appreciated is that the introduction 
of prenatal sex determination fundamentally changed the relationship between 
son preference and birth spacing.
This is because each sex-selective abortion increases the duration between 
observed births by approximately a year.%
\footnote{
This increase consists of three parts starting from the time of the 
abortion.
First, the uterus needs at least two menstrual cycles to recover 
before conception, because a less than three months space 
between abortion and conception substantially increases the likelihood 
of a spontaneous abortion \citep{zhou00b}.
Second, an average of 6 months to next conception \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.
The waiting time to conception does vary by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}
As a result, we now have a situation where families with the \emph{strongest} son 
preference likely show the \emph{longest} spacing after the birth of a daughter.
But, to further complicate matters, we may still observe short spacing 
after the birth of daughter as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.

Spacing, by itself, can therefore no longer be used to measure son preference. 
Spacing is, however, still important and useful in understanding son preference, 
but only if combined with the likelihood of observing a boy or a girl.
Short spacing after a girl is indicative of son preference if observed
births are close to the natural sex ratio, whereas long spacing is indicative
of son preference if observed births are male biased.

In this paper, I introduce and apply a novel method that directly incorporates 
the effects of sex-selective abortions on the likelihood of a son being born 
\emph{and} the duration between births.
The method can be used to analyse both situations with and without prenatal
sex selection.
Furthermore, my proposed method allow for the duration since the last 
birth to affect the decision on sex selection.
For example, as the duration from last birth becomes long enough, parents 
may reverse their decision to use prenatal sex determination,
and carry the next pregnancy to term whether male or female.
By examining whether sex selection decisions change with spacing, we can 
draw a more nuanced picture of the degree of son preference.

% [Why India?]
I apply the method to birth histories for Hindu women, using data from 
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2006. 
India is a particularly interesting case.
On one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is clearly not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period 
\citep{Yi1993,park95}.
}
On the other hand, recent research suggests that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
}
In addition, there has been substantial changes in access and legality of
prenatal sex determination in India over the period covered.
Abortion has been legal in India since 1971 and still is, but the 
first reports of on the availability of prenatal sex determination did
not appear until 1982--83 \citep{Sudha1999,bhat06,Grover2006}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
The Act is described in detail at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach 5 convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Maharashtra was the first state to pass a similar law in 1988.
}
Hence, the data make it possible to show how spacing between births and
sex ratios have changed with the introduction of prenatal sex determination,
and whether the ban affected the relationship between birth spacing and sex 
ratios.
